GPU Acceleration of the Material Point Method

### 2019-02-13 pres/pres.tex
red,sorting

\end{minipage}
\end{frame}
\begin{frame}
Nvidia Nsight\cite{NVIDIA:METRICS} now offers metrics to identify bottlenecks:
\vspace{-5}
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | l |}    \hline
    \textbf{Metric}   & \textbf{Description} \\ \hline
    VRAM SOL\%        & memory througput w.r.t. to hardware limit\\\hline
    SM SOL\%          & instruction throughput\\\hline
    L2 SOL\%          & L2-cache throughput\\\hline
    Tex SOL\%         & L1-cache throughput\\\hline
    SM Issue Util.\%  & amount of cycles an instr. was issued\\\hline
  \end{tabular}
\end{table}
\\
\vspace{5}
A simple map(\texttt{y=length(x)}) shader on 1024$\times$1024 Elements SoA vs. AoS differences:
\vspace{-15}
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
    AoS(1 instr.)         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
    SoA(1 instr.)         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\\hline
    AoS(2 instr.)         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
    SoA(2 instr.)         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
    \hline
\end{tabular}
\end{table}
$\Rightarrow$ SoA increases coalescing for non-random access.
\end{frame}
\subsection{Parallel Reduction \& Scan}
\begin{frame}
\frametitle{Parallel Reduction \& Scan}
Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ... ,a_n\right]$.\\\vspace{10}
\begin{itemize}
  \item \textbf{Parallel reduction} computes the scalar:
\begin{equation}
  r = a_0 \circ a_1 \circ ... \circ a_n.
\end{equation}
\item (Exclusive) \textbf{scan} computes the array:
\begin{equation}
  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
\end{equation}
\end{itemize}
\\\vspace{5}
Here, only shared memory approaches without (NVIDIA exclusive) warp shuffle operations.
\end{frame}
\begin{frame}
\textbf{Shared Memory Bank Conflicts}:
\\\vspace{5}
\begin{minipage}{0.94\textwidth}
\begin{minipage}{0.5\textwidth}
  \tiny
  \includesvg[svgpath =../images/,width=1.0\textwidth]{sequential}
  Sequential Addressing
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \tiny
  \includesvg[svgpath =../images/,width=1.0\textwidth]{interleaved}
  Interleaved Addressing
\end{minipage}
\end{minipage}
\\\vspace{10}
Interleaved Addressing causes bank conflicts $\Rightarrow$ padding needed.
\textbf{Sequential work}:
\end{frame}
\section{Delving Deeper: Further Opportunities}

>>>>

\end{minipage}
\end{frame}
\begin{frame}
Nvidia Nsight\cite{NVIDIA:METRICS} now offers metrics to identify bottlenecks:
\vspace{-5}
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | l |}    \hline
    \textbf{Metric}   & \textbf{Description} \\ \hline
    VRAM SOL\%        & memory througput w.r.t. to hardware limit\\\hline
    SM SOL\%          & instruction throughput\\\hline
    L2 SOL\%          & L2-cache throughput\\\hline
    Tex SOL\%         & L1-cache throughput\\\hline
    SM Issue Util.\%  & amount of cycles an instr. was issued\\\hline
  \end{tabular}
\end{table}
\\
\vspace{5}
A simple map(\texttt{y=length(x)}) shader on 1024$\times$1024 Elements SoA vs. AoS differences:
\vspace{-15}
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
    AoS(1 instr.)         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
    SoA(1 instr.)         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\\hline
    AoS(2 instr.)         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
    SoA(2 instr.)         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
    \hline
\end{tabular}
\end{table}
$\Rightarrow$ SoA increases coalescing for non-random access.
\end{frame}
\subsection{Parallel Reduction \& Scan}
\begin{frame}
\frametitle{Parallel Reduction \& Scan}
Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ... ,a_n\right]$.\\\vspace{10}
\begin{itemize}
  \item \textbf{Parallel reduction} computes the value:
\begin{equation}
  r = a_0 \circ a_1 \circ ... \circ a_n.
\end{equation}
\item (Exclusive) \textbf{scan} computes the array:
\begin{equation}
  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
\end{equation}
\end{itemize}
\\\vspace{5}
Here, only shared memory approaches without (NVIDIA exclusive) warp shuffle operations.
\end{frame}
\begin{frame}
\textbf{Shared Memory Bank Conflicts}:
\\\vspace{10}
\begin{minipage}{0.94\textwidth}
\begin{minipage}{0.5\textwidth}
  \tiny
  \includesvg[svgpath =../images/,width=1.0\textwidth]{sequential}
  Sequential Addressing
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \tiny
  \includesvg[svgpath =../images/,width=1.0\textwidth]{interleaved}
  Interleaved Addressing
\end{minipage}
\end{minipage}
\\\vspace{10}
\\
Interleaved addressing causes bank conflicts (Short Scoreboard activity) $\Rightarrow$ padding needed.
\vspace{-12}
\begin{table}
    \footnotesize
  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
    Interl. no padd.    & 305                   & -       &23.0\%          & \textbf{60.9}\%  & S. Scoreb.(17.2\%)\\\hline
    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0\%)} \\\hline
  \end{tabular}
    \caption{\footnotesize Parallel reduction on $1024\times1024$ vectors with \texttt{y=length(x)} as input.}
\end{table}
\vspace{-10}
More elements than thread group size require pyramid schemes.
\end{frame}

\textbf{Sequential work}: multiple elements per thread.\begin{itemize}
  \item Memory latency hiding (Long Scoreboard up)
  \item Higher reduction factor each dispatch $\Rightarrow$ Less global memory indirections
  \item Unrolling loops can help but adds register pressure.
\end{itemize}
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
    Sequential     & 141                   & 2.16x   &49.8\%   & 37.1\%         & S. Scoreb.(2.0\%) \\\hline\hline
    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%         & L. Scoreb.(80.1\%)\\\hline
    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%         & \textbf{L. Scoreb.(84.4\%)}\\\hline
    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%         & L. Scoreb.(76.9\%)\\
    \hline
  \end{tabular}
  \caption{\footnotesize Parallel reduction on $1024\times1024$ vectors with \texttt{y=length(x)} as input. Methods have 504, 8, 4 thread groups, respectively. A GTX970 has 13 SMs.}
\end{table}
\vspace{-10}
Scan is similar but cannot profit as much from sequential work having to keep multiple elements in register memory.
\end{frame}
\subsection{Binning \& Counting Sort: Where Are You?}
\begin{frame}
\frametitle{Binning \& Counting Sort: Where Are You?}

Grid node does not know its neighboring particles $\Rightarrow$ Binning.
\\\vspace{10}
Binning combines nicely with \textbf{Counting Sort:}\\
\begin{minipage}{0.55\textwidth}
\begin{enumerate}
  \item \textbf{Binning}: Per node counting.
  \item \textbf{Scan}: Computes new memory offset for particles.
  \item \textbf{Reordering}: Give back indexing list or do deep copy.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.40\textwidth}
  \footnotesize
 \includesvg[svgpath =../images/,width=\textwidth]{binning}
 \caption{\footnotesize Binning: Particle adds one
to counter after storing it as offset.}
 \end{minipage}
\end{frame}

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\section{Delving Deeper: Further Opportunities}
